\propertitle

### 2019-11-13 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-25T16:30:40+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

Despite remarkable advances in recording technologies, the best
experimental measurements of neuronal activity can still only record a very
small sample of neurons compared to the numbers that constitute a
functionally distinguished brain region. Many probabilistic models focus on
such samples only, somehow neglecting, in their assumptions, that the
recorded neurons are a sample from a larger network. This kind of isolation
assumptions sometimes escape attention, being subtly hidden in the
mathematics. Some probabilistic models try to take also unrecorded neurons
into account individually, and therefore become very complex\mynote{refs}.
It would be useful to have models that operate in between: addressing the
larger brain area whence the sample comes, but without adding too much
complexity and detail about it. Such intermediate models would be useful,
for example, for preliminary investigations, to help us to decide which
hypotheses to discard and which to consider for more complex and costly
studies.

In the present work we propose such an intermediate model. It answers this
question: How much can the \emph{total} activity of a large neuronal
network have been, given the observation of the activity of a very small
sample thereof? This model addresses a larger brain area, avoiding the
assumption of isolation of the sample; and by focusing on the total
activity, rather than the activity of individual neurons, it remains simple
and numerically tractable.

The proposed model is based on the straightforward combination of the
maximum-entropy method and basic sampling relations from the probability
calculus, discussed in \sect~\ref{sec:model}. The maximum-entropy or
minimum-relative-entropy method \citep{jaynes1957}[much clearer
in][]{jaynes1963}{sivia1996_r2006,hobsonetal1973,jaynes1985b,grandy1980}
has been used for different kinds of estimations of the neuronal activity
of various brain areas and about other phenomena of importance to the
neurosciences, for example gene and protein interaction \citep[for
example][]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,mackeetal2009b,tkaciketal2009,roudietal2009c,barreiroetal2010,gerwinnetal2010,mackeetal2011b,ganmoretal2011,cohenetal2011,granotatedgietal2013,mackeetal2013,tkaciketal2014b,shimazakietal2015,moraetal2015,lezonetal2006,weigtetal2009}.

@@

\mynote{Say something more about advantages of such question/answer: for
  example we can make statements about total activity of brain area even
  across recordings when sampled neurons aren't the same.}

\mynote{Add this?: from sampling theory we know that important
  features of the full network may not be visible in a sample because
  smoothed out. But using sampling theory in the inverse direction we can
  infer such full-network features from the sample.}

@@

The context we consider is as follows. During an experimental session we
have recorded the spiking activities of $n$ neurons for a certain amount of
time. These neurons are our \enquote{sample} or \enquote{subnetwork}. Their
spikes are binned into $T$ time bins and binarized to $\set{0,1}$ values in
each bin. Call $\ya_{t}$ the number of neurons that fire during time bin
$t$: this is the \emph{total activity} of the sample, or just
\enquote{activity} for short. Obviously $\ya_{t} \in\set{0,1,\dotsc,n}$; if
$\ya_{t}=0$, no neuron spikes during bin $t$; if $\ya_{t}=n$, all spike at
some point during bin $t$. For brevity, let's say \enquote{at $t$} for
\enquote{during time bin $t$}. If we divide the total activity by the
population size, $\ya/n \in \set{0,1/n,\dotsc,1}$ we have the normalized
total activity or population-averaged activity. From the activities
$\set{\ya_{t}}$ we can count how often the activity levels $\ya=0$,
$\ya=1$, and so on appeared during the recording, obtaining the
distribution of measured relative frequencies $(\yff_{\ya}) \defs \yf$. We
can also consider the sample activity at time bins \emph{outside} of the
recorded range. Such activity is unknown to us, of course.

\mynote{maybe move this paragraph to the intro?}Present-day technologies
enable the recording of neuronal activity from small brain areas \mynote{be
  more specific}. For many animal species, the neurons that are recorded
within the area are not -- and at present cannot be -- specifically chosen
from among the rest, owing to several limiting factors; for example,
limitations in how precisely electrodes are inserted or neurons are
targeted by viruses. In fact, the set of recorded neurons may even change
during very long recordings or across experimental sessions. We assume that
there's an area, comprising a network of $N$ neurons, for which we believe
that any other sample of size $n$ could have equally plausibly been
recorded instead of the sample of $n$ neurons that was actually recorded.
This is our \enquote{full network}.\mynote{how about calling it
  \enquote{the pool}?} The total activity of these $N$ neurons at $t$ is
$\yA_{t}$. The relative frequencies of the various activity levels during
the recording were $(\yFF_{\yA}) \defs \yF$. We don't know the values
$\yA_{t}$ at each $t$, or the frequency distribution $\yF$. We only know
for certain that $\yA_{t} \in \set{0, 1, \dotsc, N}$ and that
$\yA_{t} \ge \ya_{t}$ for obvious reasons. For the time being we assume
that we know $N$; in \sect*** we discuss the consequences of our lack of
precise knowledge about this number.

Our questions concern general features of the total activity $\yA$ of the
full network during and after the recording, and across sessions under the
same study conditions. For example: what was its frequency distribution
during the recording? How much does this frequency distribution change
across sessions? How much total activity should we expect at any time bin
during a recording? We cannot answer these questions with certainty; we can
only give distributions of probability or \dobs\ over their possible answers. The
approach presented here gives such probability distributions.

\mynote{Move this to intro?}We want to stress the usefulness of making a
quantified guess about the full-network activity. First of all, this guess
about a brain area seems to be the primary idea behind recording a sample
from that area. Second, it allows us to make comparisons across
experimental sessions; such comparisons would me difficult or meaningless
if made with the recorded samples, which generally comprise non-overlapping
sets of neurons and differ in size. Third, only at the full-network level
can we meaningfully assess whether correlations of some order are
informationally sufficient: as explained in
\sect~\ref{sec:marginalization}, any such assessment at the sample level
leads to conclusions opposite to the ones we think we're reaching.

@@

Regarding~\ref{item:maxent_step}, we assume that you're familiar with the
maximum-entropy method. We actually use the \emph{minimum-relative-entropy}
method \citep{hobsonetal1973}, but call it \enquote{maximum-entropy} for
brevity. It amounts to a pair of prescriptions: choose the distribution,
among those satisfying specific convex constraints, such as fixed
expectations, that has minimum relative entropy with respect to a reference
distribution, often taken to be the uniform one; and judge those
expectations to be equal to measured averages. We add two remarks about
this method that are seldom made in the literature. First remark: the
distribution $\px(\yA)$ given by this method is the zeroth-order
approximation \citep[in the sense of Laplace's
method:][\chap~4]{debruijn1958_r1961}{tierneyetal1986,strawderman2000} of
four different distributions for the full network:
\begin{itemize}
\item the most probable \emph{frequency} distribution for the total
  activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item the \emph{belief} distribution for the value of the total
  activity at any time bin among those \emph{recorded},
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item the most probable \emph{frequency} distribution for the total
  activity in a very long run of \emph{new} time bins, %$\ynuu(\yA)$,
\item the \emph{belief} distribution for the value of the total
  activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{itemize}
The maximum-entropy distribution is thus an approximation of our belief
distribution about four completely different quantities. Note that the four
distributions above numerically differ in higher-order approximations.
Second remark: the maximum-entropy method based on the Shannon entropy
implicitly makes some assumptions about the probabilities for the long-run
frequency distributions
\citep{jaynes1986d_r1996,portamana2009,portamana2017}. We say more about
these two points in \sect***.

@@

These two steps are more straightforward if instead of power moments we use
\emph{normalized factorial moments} \citep{potts1953}. The $m$th normalized
factorial moment of a distribution $\pxx(\ya)$ for the activity of the
sample neurons is defined as the average
\begin{equation}
  % \E\condt[\Bigg]{
  %   \binom{\ya}{m}\,\binom{n}{m}^{-1}}{\yH}
  % \equiv
  \sum_{\ya} %m!\,
  \binom{\ya}{m}\,  \binom{n}{m}^{-1}\, \pxx(\ya),
  \qquad 1\le m \le n
\end{equation}
% where $\px(\ya \| \yH)$ is the probability for activity $\ya$ given
% assumptions $\yH$; \enquote{$\ya$} may refer to any recorded bin or to a
% new bin.
This moment can be interpreted as the expectation of the number of distinct
$m$-tuples of simultaneously spiking neurons (within a bin's time width),
normalized by the number of distinct $m$-tuples. For example, with $m=2$,
if $na=4$ neurons spike in a network of $n=5$, we have $\tbinom{4}{2}=6$
distinct pairs of simultaneously spiking neurons, and the total number of
distinct pairs is $\tbinom{5}{2}=10$. The normalized number of spiking
pairs is therefore $6/10$. Note that the first $m$ factorial moments
provide the same information as the first $m$ power moments and vice versa:
they linearly determine each other because $\tbinom{\ya}{m}$ is a
polynomial in $\ya$ of degree $m$. Fixing the ones is therefore equivalent
to fixing the others. But the normalized factorial moments have this
extremely convenient property: \emph{the first $n$ normalized factorial
  moments for a sample and for the full network are numerically identical}:
\begin{equation}
  \label{eq:equal_m_factorialmoments}
    \sum_{\ya} %m!\,
  \binom{\ya}{m}\,  \binom{n}{m}^{-1}\, \pxx(\ya)
=    \sum_{\yA} %m!\,
  \binom{\yA}{m}\,  \binom{N}{m}^{-1}\, \pxx(\yA),
  % \E\condt[\big]{\tbinom{\ya}{m}\;\tbinom{n}{m}^{-1}}{\yH} =
  % \E\condt[\big]{\tbinom{\yA}{m}\;\tbinom{N}{m}^{-1}}{\yH},
  \qquad 1\le m \le n,
\end{equation}
where $\pxx(\yA)$ is our distribution of belief about the full-network
activity.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-13T16:55:17+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

Despite remarkable advances in recording technologies, the best
experimental measurements of instantaneous neuronal activity can still only
record a very small sample of neurons -- hundreds at most -- compared to
the numbers that constitute a functionally distinguished brain region. Many
probabilistic models focus on such samples only, somehow neglecting, in
their assumptions, that the recorded neurons are a sample from a larger
network. This kind of isolation assumptions sometimes escape attention,
being subtly hidden in the mathematics. Some probabilistic models try to
take also unrecorded neurons into account individually, and therefore
become very complex\mynote{Yasser \amp\ Vahid: refs for here?}. It would be
useful to have models that operate in between: addressing the larger brain
area whence the sample comes, but without adding too much complexity and
detail about it. Such intermediate models would be useful, for example, for
preliminary investigations, to help us to decide which hypotheses to
discard and which to consider for more complex and costly studies.

\mynote{[Yasser:] In the present work we aim to address the issue of building probability distributions over a large network using recording from subnetworks by considering the distribution of total population activity as an example. [}
In the present work we propose such an intermediate model. It answers this
question:\mynote{]} How much can the \emph{total} activity of a large neuronal
network have been, given the observation of the activity of a very small
sample thereof? This model addresses a larger brain area, avoiding the
assumption of isolation of the sample; and by focusing on the total
activity, rather than the activity of individual neurons, it remains simple
and numerically tractable.

We want to stress the usefulness of making a quantified guess about the
activity of a larger brain area. Such a guess seems indeed to be the
primary idea behind recording a sample from that area. %  Second, it allows us
% to make comparisons across experimental sessions; such comparisons would me
% difficult or meaningless if made with the recorded samples, which generally
% comprise non-overlapping sets of neurons and differ in size.
There's also an important methodological reason: the question whether
correlations of some order are informationally sufficient cannot be
properly assessed at the sample level. When the difference in size between
sample and full network is too large, correlation sufficiency for the
sample implies the \emph{lack} of correlation sufficiency for the larger
network, and vice versa. This point is explained in detail in
\sect~\ref{sec:marginalization}.

The model we propose is based on the straightforward combination of the
maximum-entropy method and basic sampling relations from the probability
calculus, discussed in \sect~\ref{sec:model}. The maximum-entropy or
minimum-relative-entropy method \citep{jaynes1957}[much clearer
in][]{jaynes1963}{sivia1996_r2006,hobsonetal1973,jaynes1985b,grandy1980}
has been used for different kinds of estimations of the neuronal activity
of various brain areas and about other phenomena of importance to the
neurosciences, for example gene and protein interaction \citep[for
example][]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,mackeetal2009b,tkaciketal2009,roudietal2009c,barreiroetal2010,gerwinnetal2010,mackeetal2011b,ganmoretal2011,cohenetal2011,granotatedgietal2013,mackeetal2013,tkaciketal2014b,shimazakietal2015,moraetal2015,lezonetal2006,weigtetal2009}.

@@

% \mynote{Say something more about advantages of such question/answer: for
%   example we can make statements about total activity of brain area even
%   across recordings when sampled neurons aren't the same.}

% \mynote{Add this?: from sampling theory we know that important
%   features of the full network may not be visible in a sample because
%   smoothed out. But using sampling theory in the inverse direction we can
%   infer such full-network features from the sample.}

@@

The context we consider is as follows. During an experimental session we
have recorded the spiking activities of $n$ neurons for a certain amount of
time. These neurons are our \enquote{sample} or \enquote{subnetwork}. Their
spikes are binned into $T$ time bins and binarized to $\set{0,1}$ values in
each bin. Call $\ya_{t}$ the number of neurons that fire during time bin
$t$: this is the \emph{total activity} of the sample, or just
\enquote{activity} for short. Obviously $\ya_{t} \in\set{0,1,\dotsc,n}$; if
$\ya_{t}=0$, no neuron spikes during bin $t$; if $\ya_{t}=n$, all spike at
some point during bin $t$. For brevity, let's say \enquote{at $t$} for
\enquote{during time bin $t$}. If we divide the total activity by the
population size we have the normalized total activity or
population-averaged activity $\ya/n$, ranging from $0$ to $1$ in $1/n$
steps. From the activities $\set{\ya_{t}}$ we can count how often the
activity levels $\ya=0$, $\ya=1$, and so on appeared during the recording,
obtaining the distribution of measured relative frequencies
$(\yff_{\ya}) \defs \yf$. We can also consider the sample activity at time
bins \emph{outside} of the recorded period. Such activity is unknown to us,
of course.

For many animal species, the neurons that are recorded within a brain area
are not specifically chosen from among the rest, owing to several limiting
factors; for example, limitations in how precisely electrodes are inserted
or neurons are targeted by viruses. The set of recorded neurons may even
change slightly across experimental sessions that are very far apart in
time. We assume that there's an area, comprising a network of $N$ neurons,
for which we believe that any other sample of size $n$ could have equally
plausibly been recorded instead of the sample of $n$ neurons that was
actually recorded. We call this larger network the \enquote{full network}.
The total activity of these $N$ neurons at $t$ is $\yA_{t}$. The relative
frequencies of the various activity levels during the recording were
$(\yFF_{\yA}) \defs \yF$. We don't know the values $\yA_{t}$ at each $t$,
or the frequency distribution $\yF$. We only know for certain that
$\yA_{t} \in \set{0, 1, \dotsc, N}$ and that $\yA_{t} \ge \ya_{t}$ for
obvious reasons. For the time being we assume that we know $N$; in
\sect~\ref{sec:N} we discuss the consequences of our lack of precise
knowledge about this number.

Our questions concern general features of the total activity $\yA$ of the
full network during and after the recording, and across sessions under the
same study conditions. For example: what was its frequency distribution
during the recording? How much does this frequency distribution change
across sessions? How much total activity should we expect at any time bin
during a recording? The approach presented here gives probability
distributions over the possible answers to these questions.

@@

Regarding~\ref{item:maxent_step}, we assume familiarity with the
maximum-entropy method. We actually use the \emph{minimum-relative-entropy}
method \citep{hobsonetal1973}, but call it \enquote{maximum-entropy} for
brevity. It amounts to a pair of prescriptions: choose the distribution,
among those satisfying specific convex constraints, such as fixed
expectations, that has minimum relative entropy with respect to a reference
distribution, often taken to be the uniform one; and judge those
expectations to be equal to measured averages. We add two remarks about
this method that are seldom made in the literature. First remark: the
distribution $\px(\yA)$ given by this method is the zeroth-order
approximation \citep[in the sense of Laplace's
method:][\chap~4]{debruijn1958_r1961}{tierneyetal1986,strawderman2000} of
four different distributions for the full network:
\begin{itemize}
\item the most probable \emph{frequency} distribution for the total
  activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item the \emph{belief} distribution for the value of the total
  activity at any time bin among those \emph{recorded},
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item the most probable \emph{frequency} distribution for the total
  activity in a very long run of \emph{new} time bins, %$\ynuu(\yA)$,
\item the \emph{belief} distribution for the value of the total
  activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{itemize}
The maximum-entropy distribution is thus an approximation of our belief
distribution about four different quantities. Note that the four
distributions above numerically differ in higher-order approximations.
Second remark: the maximum-entropy method based on the Shannon entropy
implicitly makes some assumptions about the probabilities for the long-run
frequency distributions
\citep{jaynes1986d_r1996,portamana2009,portamana2017}. We discuss these two
points further in \sect~\ref{sec:assumptions}.

@@

These two steps are more straightforward if instead of power moments we use
\emph{normalized factorial moments} \citep{potts1953}. The $m$th normalized
factorial moment of a distribution $\pxx(\ya)$ for the activity of the
sample neurons is defined as the average
\begin{equation}
  % \E\condt[\Bigg]{
  %   \binom{\ya}{m}\,\binom{n}{m}^{-1}}{\yH}
  % \equiv
  \sum_{\ya=0}^{n} %m!\,
  \displaystyle\binom{\ya}{m} \bigg/ \binom{n}{m}\; \pxx(\ya),
  \qquad 1\le m \le n
\end{equation}
% where $\px(\ya \| \yH)$ is the probability for activity $\ya$ given
% assumptions $\yH$; \enquote{$\ya$} may refer to any recorded bin or to a
% new bin.
This moment can be interpreted as the expectation of the number of distinct
$m$-tuples of simultaneously spiking neurons (within a bin's time width),
normalized by the number of distinct $m$-tuples. For example, with $m=2$,
if $a=4$ neurons spike in a network of $n=5$, we have $\tbinom{4}{2}=6$
distinct pairs of simultaneously spiking neurons, and the total number of
distinct pairs is $\tbinom{5}{2}=10$. The normalized number of spiking
pairs is therefore $6/10$. Note that the first $m$ factorial moments
provide the same information as the first $m$ power moments and vice versa:
they linearly determine each other because $\tbinom{\ya}{m}$ is a
polynomial in $\ya$ of degree $m$. Fixing the ones is therefore equivalent
to fixing the others. But the normalized factorial moments have this
extremely convenient property: \emph{the first $n$ normalized factorial
  moments for a sample and for the full network are numerically identical}:
\begin{equation}
  \label{eq:equal_m_factorialmoments}
    \sum_{\ya} %m!\,
  \binom{\ya}{m}\,  \binom{n}{m}^{-1}\, \pxx(\ya)
=    \sum_{\yA} %m!\,
  \binom{\yA}{m}\,  \binom{N}{m}^{-1}\, \pxx(\yA),
  % \E\condt[\big]{\tbinom{\ya}{m}\;\tbinom{n}{m}^{-1}}{\yH} =
  % \E\condt[\big]{\tbinom{\yA}{m}\;\tbinom{N}{m}^{-1}}{\yH},
  \qquad 1\le m \le n,
\end{equation}
where $\pxx(\yA)$ is our distribution of belief about the full-network
activity.
